Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-28 report/paper.tex
Finish draft notes on the data store

{\color{red}
This hierarchy will include some basic things that the data store understands and can
convert between (integers, floats, unbounded size integers, ISO dates, etc.)

In addition, we will define some purely semantic annotations that are commonly useful and
we will define a hierarchy for those. This represents things like addresses, postcodes, countries,
cities, etc.

Everyone can define their own schemas (like schema.org) by defining their own namespace (URL)
and providing entities with their custom namespace. The data store will simply save and return
this information - without doing any checking.}

\subsection{Content negotiation}
{\color{red}
The only bit where the data store does something clever is that it will be able to return data
in a format that the client asked for (a bit like HTTP negotiation). The client can say, ``I only
understand semantic information defined by these schema namespaces`` and the data store will do
its best to return data in the required format.

For simple types (e.g. dates) that the data store understands, it will convert data accordingly.
For other semantic information that the data store does not understand (e.g. information that
a certain cell is missing with a probability $p$), the data store will filter out the extra
information and just return the raw number.
}

>>>>

Just like the default namespace of Schema.org defines common types of entities, Wrattler
should define basic schema for the most common annotations. This includes two kinds of annotations.
First, there are primitive type annotations that are understood by the data store. Second,
there are frequent semantic annotations that are generally useful.

\paragraph{Schemas for primitive types.} The data store will store primitive values in the
data frame in a suitable native format such as floating-point number, Boolean, string or 
ISO 8601 date. However, this representation should remain internal and data store should be
able to consume and expose data in formats required by the client (this is further discussed
in Section~\ref{sec:datastore-negotiation}). This will be done through primitive type annotations.
The data store will understand those and will be able to convert between them automatically.
The primitive type annotations include:
%
\begin{itemize}
\item[--] \textbf{Numerical formats}. The data store will be able to convert between a number of common numerical
  formats such as integers, floats and unbounded size integers. In addition, values can also be
  formatted differently. Examples include the scientific notation (1.34e10) or variations in 
  decimal separators used in different cultures.

\item[--] \textbf{Dates}. The data store will be able to recognise and convert between a number of
  standard date formats (such as ISO 8601 format, dates in US and non-US formats).

\item[--] \textbf{Missing values}. The data store will record missing values internally and will be
  able to return them in a variety of formats, depending on the client request (NaN, special 
  strings such as empty or \#!NULL, etc).

\item[--] \textbf{Geo coordinates}. Finally, the list of primitive types might also include geographical
  coordinates, in which case the data store will recognize basic formats such as GPS coordinates
  and Northings and Eastings.  
\end{itemize}

\noindent
The above primitive types need to be understood by the data store, in order to enable easy
communication between clients written in a variety of programming environments that use different
primitive types. The key idea is that each client will only need to handle, for example, one format 
for dates. The data store will be able to communicate using this format, even when the value was
written in another format by another client.

In addition, Wrattler should define a set of annotations that can be used by a wide range of
standard data sets. It is useful to provide a basic schema for such common annotations as this
will enable a range of AI assistants and other tools to exchange information without first having
to define a common custom schema. Those should include annotations for places (addresses, 
countries, cities, postcodes), physical units (meters, killograms, etc.), contact information
(e-mails, phone numbers) and similar. We will investigate whether Schema.org or other existing
standards developed by the semantic web community already cover those areas.

\subsection{Content negotiation}
\label{sec:datastore-negotiation}

As noted earlier, the data store will understand a range of primitive data types and will be 
able to convert between them. This will be done using a content negotiation protocol,
similar to the Accept/Content-Type header pair in the HTTP protocol. Rather than using 
content type as in HTTP, the negotiation will be based on the schema. For example, a client may
specify that it accepts data with a certain namespace or that it recognizes a specified list of
types.

For primitive types that the data store understands, it will convert data according to the
request made by the client. For example, a date can be automatically converted between 
mm/dd and dd/mm formats. For additional semantic annotations that the data store does not
understand, it will simply drop the extra information not required by the client.

This makes it possible to exchange data efficiently. For example, a data frame where each cell
is annotated with a probabilistically inferred type will be significantly larger than a data
frame contianing just raw numbers. The content negotiation protocol allows the client to 
request a data frame with just raw numbers, significantly reducing the size.
